---
title: "Titanic Survival Analysis"
author: "Jacob Salway"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Titanic 

In this report I will attempt to predict the survival of passengers on the Titanic.

# Loading and cleaning

```{r, warning=F, message=F}
library(ggplot2)
library(dplyr)
library(janitor)

# load data
test <- read.csv('data/test.csv')
train <- read.csv('data/train.csv')

# flag so we can separate later
test$source <- "test"
train$source <- "train"

# combine train and test data
titanic <- bind_rows(train, test)

# lowercase names
titanic <- clean_names(titanic)

# quick data summary
str(titanic)
summary(titanic)
```

Missing data:

```{r}

colSums(is.na(titanic))
colSums(titanic == "")

# impute mean for missing age values
# titanic$age[is.na(titanic$age)] <- mean(titanic$age, na.rm=T)

# sample from existing data for missing embarked values
freq = data.frame(table(titanic$embarked))
freq <- freq[freq$Var1 != "",]
freq$prob <- freq$Freq / sum(freq$Freq)
freq$Var1 <- as.character(freq$Var1) # otherwise it samples with integers
size = nrow(titanic[titanic$embarked == "",])
titanic$embarked[titanic$embarked == ""] <- sample(freq$Var1, prob=freq$prob, replace=T, size=size)

# verify
colSums(is.na(titanic))
colSums(titanic == "")
```

Reclassifying features:

```{r}
# convert to factors
cols = c("survived", "pclass", "embarked", "source")
for (i in cols) {
  titanic[,i] <- as.factor(titanic[,i])
}

# family size
titanic$family_size <- titanic$sib_sp + titanic$parch + 1

# new structure
str(titanic)
```

# Analysis

Because we've cleaned the data, it's time to separate it back into training and test data.

```{r}
train <- titanic[titanic$source == "train",]
test <- titanic[titanic$source == "test",]
```

Let's start with some general demographic information.

```{r}
# setting up some charting presets
age_x <- scale_x_continuous(breaks=seq(0, 100, 10))
prop_y <- scale_y_continuous()

# age and gender distribution
t = titanic[!is.na(titanic$age),]

ggplot(t, aes(x=age, fill=sex)) +
  geom_histogram(binwidth=3) +
  labs(x="Age", y="Count", title="Age and gender") +
  scale_x_continuous(breaks=seq(0, 100, 10))

# gender proportions
t = titanic %>%
  filter(!is.na(age)) %>%
  mutate(age=round(age/3)*3) %>%
  group_by(age, sex) %>%
  summarise(n=n()) %>%
  mutate(prop=n/sum(n))

ggplot(t, aes(x=age, y=prop)) +
  geom_bar(aes(fill=sex), stat="identity") +
  labs(x="Age", y="Proportion", title="Gender proportions") +
  scale_x_continuous(breaks=seq(0, 100, 10)) +
  scale_y_continuous(breaks=seq(0, 1, 0.25), labels=scales::percent) +
  geom_hline(yintercept=0.5)
```

Now let's try to visually see if any variables had a major impact on whether a person survived.

## Variable analysis

Let's start by defining a group proportion function.

```{r}
calc_prop <- function(x, y) {
  t = table(x, y)
  for (i in 1:dim(t)[1]) {
    t[i,] <- round(t[i,]/sum(t[i,])*100)/100
  }
  return(t)
}
```

### Sex

```{r}
t = calc_prop(train$sex, train$survived)
t

ggplot(train, aes(x=sex, fill=survived)) +
  geom_bar(position="fill")
```

### Age

```{r}
t = train %>%
  filter(!is.na(age))

ggplot(t, aes(x=age, fill=survived)) +
  geom_histogram(binwidth=3) +
  labs(x="Age", y="Count", title="Age and survival") +
  age_x

ggplot(t, aes(x=age, fill=survived)) +
  geom_histogram(binwidth=3, position="fill") +
  labs(x="Age", y="Proportion", title="Age and survival proportions") +
  age_x + prop_y
```

### Embarked

```{r}
t = calc_prop(train$embarked, train$survived)
t

ggplot(train, aes(x=embarked, fill=survived)) +
  geom_bar(position="fill")
```

### Passenger class

```{r}
t = calc_prop(train$pclass, train$survived)
t

ggplot(train, aes(x=pclass, fill=survived)) +
  geom_bar(position="fill")
```

### Fare

```{r}
ggplot(train, aes(x=fare, fill=survived)) +
  geom_histogram(binwidth=30, position="fill") +
  scale_x_continuous(breaks=seq(0, 1000, 100))
```

### Family size

```{r}
t = calc_prop(train$family_size, train$survived)
t

ggplot(train, aes(x=family_size, fill=survived)) +
  geom_bar(position="fill") +
  scale_x_continuous(breaks=seq(1, 20, 1))
```

### Conclusion

# Prediction